A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-05-29 papers/water-SAFT/paper.tex
Reran two rods simulations. Added files for single rod density plot.  Changed some wording in conclusion and introduction. Added small section for contact-value theorem. Tested everything on quipu.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT-VR). Using this approximation with an additional length scale parameter, we are able to calculate the vapor pressure, saturated liquid density, surface tension, and enthalpy of water, as well as the number of hydrogen bonds and density oscillations formed near hydrophobic rods and spheres immersed in water. 
For two rods and four rods(arranged as the corners of a square) in water, our functional predicts
a transition from a vapor state to a liquid state in the area between
the rods that agrees with clasical theory.   We find that in the vapor state, the force per length can be
approximated classically and is equal to 144~mN/m, or twice the bulk surface 
tension even for nm scale rods, and  our hydrophobic spheres agree
reasonably well with SPC/E simulations.
This functional is computationally efficient and provides a foundation for 
future work describing complex interactions in aqueous solution on an
atomic level.}
\tableofcontents
\end{abstract}
\maketitle

@@

Due to its large dielectric constant, small molecular size and strong
hydrogen-bonding network, water as ``the universal solvent'' is as
interesting as its anomalous thermodynamic properties.  \fixme{what
  the heck?}

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\fixme{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\fixme{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by \fixme{how much? significant factor, I
  think}, and are unsuited for studying systems with molecular-scale
inhomogeneities.

%% A recent functional presented by Lischner and Arias~\cite{lischner2010classical}
%% describe water as non-interacting molecules in a an
%% effective external potential.  This functional contains the exact
%% entropy for no-interacting molecules and, among other things, can 
%% predict the nonlinear dielectric response in water, but it must be paramaterized
%% for different temperatures.

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact density
theorem, which states that the contact density at a hard surface $n_c$
is proportional to the pressure on that surface. This
also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes: 
\begin{equation}\label{contactvaluethm}
  F = n_c k_BT V\:
\end{equation}
These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}.  A common
failing of SAFT-based functionals (see, for instance
references\cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}) is to treat other terms in the
functional locally, resulting in a functional that is suitable only
for studying the liquid-vapor interface.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii (0.3~nm, 0.5~nm, and 1~nm) which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. We use the
contact value theorem to derive the contact density at the water-rod
interface.  We start with the thermodyamic expression relating
pressure to the partial derivative of free energy with respect the
radius of a cylinder. Using the contact value theorem and sustituting in
the free energy gradient for the pressure, we can derive values for
the contact density from data used in Figure
\ref{fig:energy-vs-diameter}.  As the radius of the rods approch zero we
expect the contact density to approach the bulk density, which our
functional does.  This self consistency shows that our functional is
thermodynamically correct.

@@

The free energy per area for water surrounding a hydrophobic rod shows the
expected behavior with an asymptote equal to the surface tension.
Based on this result, we expect to be
able to classically describe rods with a radius of about 0.5~nm or larger. 
For two hydrophobic rods surrounded by water, we see a transition from a vapor-filled
stadium shape to two rods with liquid between them. A
classical treatment of the critical separation for this transition
works well for rods larger than 0.5~nm.  Further evidence for the
accuracy of this classical treatment is the calculation of the force on the
stadium shape. The force per length is constant and equal to twice the
surface tension. A simulations of hydrophobic spheres agrees with
SPC/E calculations and a system with four hydrophobic rods exhibited
similar behavior as the two rod.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
{We develop a classical density functional theory for water by combining the
fundamental-measure theory functional--which describes hard sphere fluids--with
attractive interactions based on the Statistical Associating Fluid Theory
(SAFT-VR). Using this approximation with an additional length scale parameter, we are able to calculate the vapor pressure, saturated liquid density, surface tension, and enthalpy of water, as well as the number of hydrogen bonds and density oscillations formed near hydrophobic rods and spheres immersed in water. 
For two rods and four rods(arranged as the corners of a square) in water, our functional predicts
a transition from a vapor state to a liquid state in the area between
the rods that agrees with clasical theory.   We find that in the vapor state, the force per length can be
approximated classically and is equal to 144~mN/m, or twice the bulk surface 
tension even for nm scale rods, and  our hydrophobic spheres agree
reasonably well with SPC/E simulations.
This functional is computationally efficient and provides a foundation for 
future work describing complex interactions in aqueous solution on an
atomic level.}
\end{abstract}
\maketitle

@@

As common as water is, its large dielectric constant, small
molecular size, strong hydrogen-bonding network and anomolous
thermodynamic properties make ``the universal solvent'' an interesting
subject of study.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\fixme{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\fixme{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by a factor of about 1.5, and are unsuited for studying
systems with molecular-scale inhomogeneities.

\subsection{Contact-value theorem}

The contact value theorem states that the pressure of a fluid at a
hard wall interface is proportional to the contact density of the fluid,
$n_c$.  Specifically,
\begin{equation}\label{eq:contact}
  P = n_ck_BT
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ is the temperature of the
fluid. This leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes: 
\begin{equation}\label{contactvaluethm}
  F = n_c k_BT V\:
\end{equation}

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact value
theorem and its limiting case of small solutes (Equation
\ref{contactvaluethm}). These properties are retained by the total
functional, provided the remaining terms are \emph{purely nonlocal}.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density profiles for
different radii rods which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. We use the
contact value theorem to derive the contact density at the water-rod
interface.  We start with the thermodyamic expression relating
pressure to the partial derivative of free energy with respect the
radius of a cylinder. Using the contact value theorem and sustituting in
the free energy gradient for the pressure, we can derive values for
the contact density from data used in Figure
\ref{fig:energy-vs-diameter}.  As the radius of the rods approch zero we
expect the contact density to approach the bulk density, which our
functional does.  This self consistency shows that our functional is
thermodynamically correct.

@@

The free energy per area for water surrounding a hydrophobic rod shows the
expected behavior with an asymptote equal to the surface tension.
Based on this result, we expect to be
able to classically describe a single rod with a radius of about 0.5~nm or larger. 
For two hydrophobic rods surrounded by water, we see a transition from a vapor-filled
stadium shape to two rods with liquid between them. A
classical treatment of the critical separation for this transition
works well for pairs of rods with radii larger than 0.5~nm.  Further evidence for the
accuracy of this classical treatment is the calculation of the force on the
stadium shape. The force per length is constant and equal to twice the
surface tension. A similar system with four hydrophobic rods exhibited the same
behavior while simulations of hydrophobic spheres agrees with
SPC/E calculations.
